Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2018-07-17 workflow/workflow.Rmd
Edit to text

scRNA-seq low-level analysis workflows have already been developed, with useful methods for quality control (QC), exploratory data analysis (EDA), pre-processing, normalization, and visualization. The workflow described in [@Lun2016] and the package `scater` [@McCarthy2017] are such examples based on open-source R software packages from the Bioconductor Project [@Huber2015]. In these workflows, single-cell expression data are organized in objects of the `SCESet` class allowing integrated analysis. However, these workflows are mostly used to prepare the data for further downstream analysis and do not focus on steps such as cell clustering and lineage inference.

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages. Throughout the workflow, we use a single `SingleCellExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment.

@@

EP: I have only updated the code, not the text. The text needs adaptation, especially future directions and references to `SummarizedExperiment` which should be `SingleCellExperiment`.
FP: TODO, fix ref.

@@

The scRNA-seq dataset we use as a case study was generated to study the differentitation of HBC stem cells into different cell types present in the olfactory epithelium. To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing. The expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it. Cells were then assigned to different lineages using a statistical analysis pipeline analogous to that in the present workflow. Finally, results were validated experimentally using in vivo lineage tracing. Details on data generation and statistical methods are available in [@Fletcher2017; @Risso2017; @Street2017].

@@

Note that in order to successfully run the workflow, we need the devel versions
of the Bioconductor packages `scone (>=1.5.0)`, `zinbwave (>=1.3.2)`, and `clusterExperiment (>=2.1.3)`. We recommend running Bioconductor 3.8 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).

EP: Should update these.

DR: Done, but let's double-check again before the final submission.

@@

To give an idea to the users of the time needed to run the workflow, function `system.time` was used to report computation times for the time consuming functions. Computations were performed with 2 cores on a MacBook Pro (early 2015) with a 2.7 GHz Intel Core i5 processor and 8 GB of RAM. The `BiocParallel` package is used to allow for parallel computing in `zinbwave` function. Users with a different operating system may change the package used for parallel computing and the `NCORES` variable below.

@@

EP: Note that clusterExperiment uses `parallel` package, so might should update this.

@@

EP: There's code in SingleCellExperiment for spike-ins. We should use it, though I'm not familiar with it. 
FP: Davide, what is the standard way to use SingleCellExperiment for spike-ins?

@@

Throughout the workflow, we use the class `SingleCellExperiment` to keep track of the counts and their associated metadata within a single object. The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to any cluster in the original publication.

@@

# Symbol for cells not assigned to a lineage in original data
metadata$publishedClusters[is.na(metadata$publishedClusters)] <- -2

dataObj <- SingleCellExperiment(assays = list(counts = E),
                           colData = metadata)
dataObj
```

Using the Bioconductor R package `scone`, we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes. See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.

@@

Finally, for computational efficiency, we retain only the 1,000 most variable genes. This seems to be a reasonnable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]). In general, however, we recommend care in selecting a gene filtering scheme, as an appropriate choice is dataset-dependent.

```{r}
# # Filtering to top 1,000 most variable genes
vars <- rowVars(log1p(counts(dataObj)))
names(vars) <- rownames(dataObj)
vars <- sort(vars, decreasing = TRUE)
# dataObj <- dataObj[names(vars)[1:1000],]

#alternative script:
dataObj<-makeFilterStats(dataObj,filterStats="var", transFun = log1p)
dataObj<-filterData(dataObj,percentile=1000,filterStats="var")

length(intersect(rownames(dataObj), names(vars)[1:1000]))
setdiff(rownames(dataObj), names(vars)[1:1000])
```

EP: Added existing code in the clusterExperiment package that does this. Also will mean that any future steps in package that need variance for the dataObj will have variance per gene stored.  
FP: great. I added the filterStats argument to filterData function. I get an error otherwise.

FP: We get a different list of most variable genes here.

DR: Fixed, the new code was computing the variance in linear space instead of using the log.

DR: Remember to remove the old code and the length() and setdiff() from above.

@@

The function `zinbwave` returns a `SummarizedExperiment` object that includes normalized expression measures, defined as deviance residuals from the fit of the ZINB-WaVE model with user-specified gene- and cell-level covariates. Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots). Note that, in this case, the low-dimensional matrix `W` is not included in the computation of residuals to avoid the removal of the biological signal of interest.

@@

FP: These values are slightly different from what

@@

EP: This saves a lot of trouble with the plotHeatmap, etc.
FP: Ok.

EP: I've changed this code from my previous ad hoc to not delete the previous, but just reorder them. Again, (it might make sense for me to add option to different assays in clusterExperiment, but I'm not going to do that right now.)

EP: Should probably address these normalized values, especially to more strongly make clear that shouldn't use these for clustering, just for visualization. The way its stated it sounds like dimensionality reduction is an "also", when in fact it is the main point and the normalized values are the afterthought. Might consider reordering the sections to more clearly demonstrate this and put dimensionality reduction first.  
FP: Agree, let's keep it in mind for once the workflow runs smoothly.

DR: Agree, I'm not even sure normalized values are that useful for visualization... a simple scaling by total counts would probably achieve the same... but since we have them...

@@

The `zinbwave` function can also be used to perform dimensionality reduction, where, in this workflow, the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`. The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance. To verify that `W` indeed captures the biological signal of interest, we display the MDS results in a scatterplot with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).

```{r mdsW, fig.cap="ZINB-WaVE: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
W <- reducedDim(dataObj)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(publishedClusters)], main = "",
     pch = 20, xlab = "Component 1", ylab = "Component 2")
legend(x = "topleft", legend = unique(names(col_clus)), cex = .5, fill = unique(col_clus), title = "Sample")
```

@@

The next step of the workflow is to cluster the cells according to the low-dimensional matrix `W` computed in the previous step. We use the resampling-based sequential ensemble clustering (RSEC) framework implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`. Specifically, given a set of user-supplied base clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings, with the option of resampling cells and using a sequential tight clustering procedure as in [@Tseng2005]. A consensus clustering is obtained based on the levels of co-clustering of samples across the candidate clusterings. The consensus clustering is further condensed by merging similar clusters, which is done by creating a hierarchy of clusters, working up the tree, and testing for differential expression between sister nodes, with nodes of insufficient DE collapsed. As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.

Note that the defaults in `RSEC` are designed for input data that are the actual (normalized) counts. Here, we are applying `RSEC` instead to the low-dimensional `W` matrix from ZINB-WaVE, for which we make a separate `SummarizedExperiment` object. For this reason, we choose to not use certain options in `RSEC`. In particular, we do not use the default dimensionality reduction step, since our input `W` is already in a space of reduced dimension. Specifically, `RSEC` offers a dimensionality reduction option for the input to both the clustering routines (`reduceMethod`) and the construction of the hiearchy between the clusters (`dendroReduce`). We also skip the option to merge our clusters based on the amount of differential gene expression between clusters.

FP: As the name/type "zinbwave50" is then needed in the RSEC function, we might need to keep this line.
EP: I just called it "zinbwave50" for convenience. I've changed future code to just use "zinbwave". Though as a suggestion, you might consider for the zinbwave function to add the value of K to the name automatically, since the dimensions are not nested within each other, like pca, so you might run multiple versions of zinbwave and want to compare different K.

EP: Note that we need to see whether clusterExperiment retains all of the different assays or not. It should in principle, but its not something I've tested. Also, I purposefully didn't give the parameter `nReducedDims` in RSEC because I think it should automatically use the max saved in object if not designated. If we're not getting the same results, we can add in `nReducedDims=50` and see; if that changes anything, let me know. However, I would focus on the question of different W values.

FP: I am not sure it makes sense to perform the DE on the normalized values. Intuitively, I would say no, but not sure. We can still show that it is a possibility and write in the text that in this case, it might not make sense. We are working on a DE method for scRNA-seq where we use observational weights computed from the posterior probabilities of the ZINB-WaVE model in bulk RNA tools like edgeR and DESeq2. I think, eventually, it would be nice to be able to use that. Maybe, not for the next next version of the workflow.  
EP: I'd note that we use the normalized values for finding those DE with respsect to lineage. But I don't think its necessary to put it in here, but maybe in future directions, where you discuss that functionality, we could mention that it would allow the usage of mergeClusters. Also would be a good time to point out that DE on normalized values probably not a great idea either, if you don't think so. I also think the reason for skipping mergeClusters should be stated in the above text, rather than just saying we are skipping it.

DR: A lot of these comments are now outdated, as the new version of clusterExperiment fully support multiple assays. Similarly, the weights are now a reality and also supported by clusterExperiment. We should probably add a section about it?

@@

FP: We get more clusters for the makeConsensus method than before. Maybe because the most variable genes are different?

DR: I fixed the most variable genes. Is there a way to check the older results? Where do I find the saved objects?

@@

FP: Ordering does not seem right.  
EP: I get a very different plotCoClustering plot than what you sent me. See email.

@@

FP: we do not get exactly the same number of cells in each cluster. It looks similar though.

DR: it's pretty similar, but we still get 7 clusters that look more or less the same as before.

@@

The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figure \@ref(fig:addPublishedClusters)), the main difference being that several of the published clusters were merged here into single clusters. This discrepancy is likely caused by the fact that we started with the top 1,000 genes, which might not be enough to discriminate between closely related clusters.

@@

FP: compared to the previous version, it seems that there is a lot of black. Would the black coming from the contour of the rectangles in the bars?
EP: I'm not sure what you mean about a lot of black on the barplot. Do you mean the unassigned samples? The colors have changed because the default set of colors was updated.

DR: Now it looks pretty much the same.

DR: Better to use `plotClustersTable` here? See below.

@@

FP: publishedClusters does not appear in the bars.
EP: I forgot to add it to the `whichClusters` argument.

@@

```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
plotReducedDims(dataObj,whichCluster="primary",reducedDim="zinbwave",pch=20,
	xlab = "Component1", ylab = "Component2",legendTitle="Sample",main="",
	plotUnassigned=FALSE
)

# ###test
# col<-convertClusterLegend(dataObj,whichCluster="primary",output="matrixColors")[,1]
# col[col=="white"]<-"grey"
# W<-reducedDim(dataObj,"zinbwave")
# plot(-W[,1:2],col=col,pch=20)
```

FP: I get `"reduceMethod" is not a graphical parameter`. I think this plot is different from the previous version.

EP: I had wrong name for argument which has changed (should have been `reducedDim`), so it went with default which is PCA on the assay(dataObj) slot -- i.e. not using zinbwave. Now it gives the same plot as my test, where I directly plotted the results of `reducedDim(dataObj,"zinbwave")`. However, its still not exactly the same as previous rendition online. First it looks like the dimensions got flipped, i.e. multiplied by -1. Furthermore, it doesn't look exactly the same (e.g. outlying points not the same), even if I color in the unassigned so we can see them. Is zinbwave giving back the exact same W as before? I've checked that the W in the `dataObj_afterZinbwave` and W I plotted above from `dataObj_afterRSEC` are the same so its not clusterExperiment messing something up. If so, this could be why we are getting slightly different clustering results.

DR: It is possible that zinbwave does not return the same exact W. There is no guarantee of a global maximum, but only local. The plot looks more or less the same, but just flipped.

@@

We now demonstrate how to use the R software package `slingshot` to infer branching cell lineages and order cells by developmental progression along each lineage. The method, proposed in [@Street2017], comprises two main steps: (1) The inference of the global lineage structure (i.e., the number of lineages and where they branch) using a minimum spanning tree (MST) on the clusters identified above by `RSEC` and (2) the inference of cell pseudotime variables along each lineage using a novel method of simultaneous principal curves. The approach in (1) allows the identification of any number of novel lineages, while also accommodating the use of domain-specific knowledge to supervise parts of the tree (e.g., known terminal states); the approach in (2) yields robust pseudotimes for smooth, branching lineages.

The two steps of the Slingshot algorithm are implemented in the functions `getLineages` and `getCurves`, respectively. The first takes as input a low-dimensional representation of the cells and a vector of cluster labels. It fits an MST to the clusters and identifies lineages as paths through this tree. The output of `getLineages` is an object of class `SlingshotDataSet` containing all the information used to fit the tree and identify lineages. The function `getCurves` then takes this object as input and fits simultaneous principal curves to the identified lineages. These functions can be run separately, as below, or jointly by the wrapper function `slingshot`.

@@

Cluster name | Description | Color | Correspondence
-------------|-------------|-------| ----------
c1 | HBC | blue | original 1, 5
c2 | GBC | green | original 2, 3, 11
c3 | mSUS | red | original 4, 7
c4 | Contaminants | orange | original -2 
c5 | mOSN | purple | original 9, 12
c6 | Immature Neuron | brown | original 10, 14
c7 | MV | light blue | original 15

Cells in cluster `c4` have a cluster label of `-2` in the original published clustering, meaning that they were not assigned to any cluster. These cells were actually identified as non-sensory contaminants, as they overexpress gene `Reg3g` (see Figure S1 from [@Fletcher2017] and Figure \@ref(fig:boxplotReg3g)), and were removed from the original published clustering. While it is reassuring that our workflow clustered these cells separately, with no influence on the clustering of the other cells, we removed cluster `c4` to infer lineages and pseudotimes, as cells in this cluster do not participate in the cell differentiation process. Note that, out of the 77 cells overexpressing `Reg3g`, 11 are captured in cluster `c4` and 21 are unclustered in our workflow's clustering (see Figure \@ref(fig:boxplotReg3g)). However, we retain the remaining 45 cells to infer lineages as they did not seem to influence the clustering.

DR: This has changed (slightly). This is the table that I get currently.

@@

```{r boxplotReg3g, fig.width=6, fig.height=4, fig.cap="RSEC: Boxplots of the log count of gene Reg3g stratified by cluster."}
plotFeatureBoxplot(dataObj, whichCluster="primary", feature="Reg3g",ylab = "Reg3g count values", cex.axis = .8, cex.lab = .8,assay=log1p(assays(dataObj)[["counts"]]))
```

FP: this plot is fairly different from the previous version. We should see for one cluster that the gene expression for Reg3g is much high that for the other clusters.

EP: I think this plot showed the normalized values (saved in assay(dataObj)). But I see that the previous plot was on the counts. Is that better than normalized value? I have added (just to this plot for now), an option to plot a specific given matrix (or a character value defining an assay, but since we want log-transform of the counts, which is not saved in the main assay, putting in "counts" doesn't give plot we want). But it's now cluster c1, not c4 that has this property. It might be useful to take the clustering from the old RSEC data object and directly compare the clusterings. Regardless, none of the previous commands are going to get the right results going forward since the clusters don't have the same name. See note on plotReducedDims command where I think the input `W` has changed.

DR: I think we should re-write this section. With the current version of the package and parameters, we don't get the contaminant cluster anymore, but most of the contaminant cells are now in c1. However c1 also includes HBC cells, so we should not remove it from the slingshot analysis.

@@

```{r}
pseudoCe <- dataObj[,!primaryClusterNamed(dataObj) %in% c("-1")]
X <- reducedDim(pseudoCe,type="zinbwave")
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
reducedDim(pseudoCe, "MDS") <- mds$points
pseudoCe <- slingshot(pseudoCe, reducedDim = "MDS", start.clus = "c1")
```
EP: See my email. I don't get this subsetting error on my machine or on TravisCI. I do get on SCF Servers which are running R3.4.1 and the older version of bioconductor. Can you check this on your machine?

FP: I agree. I get an error though when I run line 501:
`Error in .BasicFunsList[[f]] : no such index at level 1`.

FP: I stooped here.

DR: Above, I use the syntax to work directly on the ClusterExperiment object, but if we want to be able to plot the resulting lineages, we need the matrix method.

Before fitting the simultaneous principal curves, we examine the global structure of the lineages by plotting the MST on the clusters. This shows that our implementation has recovered the lineages found in the published work (Figure \@ref(fig:tree)). The `slingshot` package also includes functionality for 3-dimensional visualization as in Figure \@ref(fig:stemcelldiff), using the `plot3d` function from the package `rgl`.

```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with connecting lines between cluster centers representing the inferred global lineage structure."}
colorCl<-convertClusterLegend(pseudoCe,whichCluster="primary",output="matrixColors")[,1]
lineages <- getLineages(mds$points, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1")
pairs(lineages, type="lineages", col = colorCl)
```

DR: For some reason, this plot does not work anymore, but the lineages that I get are correct!

Having found the global lineage structure, we now construct a set of smooth, branching curves in order to infer the pseudotime variables. Simultaneous principal curves are constructed from the individual cells along each lineage, rather than the cell clusters. This makes them more stable and better suited for assigning cells to lineages. The final curves are shown in Figure \@ref(fig:curves).

```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with smooth curves representing each inferred lineage."}
lineages <- getCurves(lineages)
pairs(lineages, type="curves", col = colorCl)
```

@@

In the workflow, we recover a reasonable ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. `getLineages` has the option for the user to provide known end cluster(s). Here is the code to use `slingshot` in a supervised setting, where we know that clusters `c3` and `c7` represent terminal cell fates.

```{r slingshotsupervised, eval=FALSE}
lineages <- getLineages(X, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1",
                         end.clus = c("c3", "c7"))
lineages <- getCurves(lineages)
pairs(lineages, type="curves", col = colorCl)
pairs(lineages, type="lineages", col = colorCl, 
      show.constraints = TRUE)

lineages
```

@@

EP: Since changed NormalizedValues to the main assay, this cleans up the code above and below. I've used the `transformData` command, which will not do anything if isCount=FALSE above, but will convert to right if we change isCount=TRUE. It's better to use that command than `assay` for that reason. Might should add comment in code or text explaining this.

@@

EP: note the above does heirarchcial clustering of the top genes. If want them to stay in order given, should give `clusterFeaturesData=topgenes` (I think that is the argument)

EP: The above code should take care of the ordering and NA values. Note that subseting a ClusterExperiment object is suppose to also subset the orderSamples value correctly. Let me know if that is not the case.

@@

In an effort to improve scRNA-seq data analysis workflows, we are currently exploring a variety of applications and extensions of our ZINB-WaVE model. In particular, we are developing a method to impute counts for dropouts; the imputed counts could be used in subsequent steps of the workflow, including dimensionality reduction, clustering, and cell lineage inference. In addition, we are extending ZINB-WaVE to identify differentially expressed genes, both in terms of the negative binomial mean and the zero inflation probability, reflecting, respectively, gradual DE and on/off DE patterns. We are also developing a method to identify genes that are DE either within or between lineages inferred from Slingshot.

Finally, a new S4 class called `SingleCellExperiment` is currently under development (https://github.com/drisso/SingleCellExperiment). This new class is essentially a `SummarizedExperiment` class with a couple of additional slots, the most important of which is `reducedDims`, which, much like the `assays` slot of `SummarizedExperiment`, can contain one or more matrices of reduced dimension. This new `SingleCellExperiment` class would be a valuable addition to the workflow, as we could store in a single object the raw counts as well as the low-dimensional matrix created by the ZINB-WaVE dimensionality reduction step. Once the implementation of this class is stable, we would like to incorporate it to the workflow.

EP:  Obviously should replace this text. Might want to mention ability of class to store data matrices not in memory, though not currently implemented, which is the next major change we need to do here.

@@

>>>>

scRNA-seq low-level analysis workflows have already been developed, with useful methods for quality control (QC), exploratory data analysis (EDA), pre-processing, normalization, and visualization. The workflow described in [@Lun2016] and the package `scater` [@McCarthy2017] are such examples based on open-source R software packages from the Bioconductor Project [@Huber2015]. In these workflows, single-cell expression data are organized in objects of the `SingleCellExperiment` class allowing integrated analysis. However, these workflows are mostly used to prepare the data for further downstream analysis and do not focus on steps such as cell clustering and lineage inference.

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` Bioconductor package; and (4) DE analysis along lineages. Throughout the workflow, we use a single `SingleCellExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment. The use of the `SingleCellExperiment` class allows us to seamlessly transition between the steps of this workflow, as well as to integrate this workflow with the upstream QC and EDA steps provided in [@Lun2016].

@@

@@

The scRNA-seq dataset we use as a case study was generated to study the differentitation of HBC stem cells into different cell types present in the olfactory epithelium. To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing. The expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it. Cells were then assigned to different lineages using a statistical analysis pipeline analogous to that in the present workflow. Finally, results were validated experimentally using in vivo lineage tracing. Details on data generation and statistical methods are available in [@Fletcher2017; @Risso2017; @Risso2018; @Street2017].

@@

Note that in order to successfully run the workflow, we need the following versions
of the Bioconductor packages `scone (>=1.5.0)`, `zinbwave (>=1.3.2)`, and `clusterExperiment (>=2.1.3)`. We recommend running Bioconductor 3.8 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).

@@

To give an idea to the users of the time needed to run the workflow, function `system.time` was used to report computation times for the time consuming functions. Computations were performed with 2 cores on a iMac with a 4 GHz Intel Core i7 processor and 32 GB of RAM. The `BiocParallel` package is used to allow for parallel computing in the `zinbwave` function. Users with a different operating system may change the package used for parallel computing and the `NCORES` variable below.

@@

@@

DR: Consider keeping the spike-ins in the object to show the ability of SCE to use spike-ins... but we don't use them, so not sure how important it is.

@@

The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to any cluster in the original publication.

@@

# Symbol for cells not assigned to a lineage in original data
metadata$publishedClusters[is.na(metadata$publishedClusters)] <- -2
```

Throughout the workflow, we store the data in an object of class `SingleCellExperiment`. The `SingleCellExperiment` class [@SingleCellExperiment] is an extension of the `SummarizedExperiment` class [@SummarizedExperiment] specifically designed for single-cell data. In addition to keeping the expression values and their associated metadata within a single object, `SingleCellExperiment` allows the user to store normalization factors and the results of dimensionality reduction methods (such as `zinbwave`) within the object. Throughout the workflow, we will highlight when this special features of `SingleCellExperiment` are used.

```{r}
dataObj <- SingleCellExperiment(assays = list(counts = E),
                           colData = metadata)
dataObj
```

Using the Bioconductor R package `scone` [@Cole235382], we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes. See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.

@@

Finally, for computational efficiency, we retain only the 1,000 most variable genes. This seems to be a reasonable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]). In general, however, we recommend care in selecting a gene filtering scheme, as an appropriate choice is dataset-dependent.

To compute the gene-wise variances and keep only the most variable genes, we use two functions from the `clusterExperiment` package. In particular, `makeFilterStats` will compute the variance of the gene expression data with a given transformation applied to the counts (in this case `log(x+1)`). It will then store the variance values in the `rowData` slot of the object. `filterData` will use the computed values to keep only the desired number of genes.

```{r}
dataObj <- makeFilterStats(dataObj, filterStats="var", transFun = log1p)
dataObj <- filterData(dataObj, percentile=1000, filterStats="var")
rowData(dataObj)
```

@@

The function `zinbwave` returns a `SingleCellExperiment` object that includes the normalized expression measures (`notmalizedValues`), defined as deviance residuals from the fit of the ZINB-WaVE model with user-specified gene- and cell-level covariates. Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots). Note that, in this case, the low-dimensional matrix `W` is not included in the computation of residuals to avoid the removal of the biological signal of interest.

@@

@@

In addition to the normalized values, the `zinbwave` function added several new components to the object: the assay `residuals` contains the residuals of the model that can be used to check the goodness-of-fit of the model. The `weights` assay contains observational weights that can be used to downweight the effect of zero counts on differential expression (see [@van2018observation] and the `zinbwave` vignette for more details).

@@

Perhaps the most important component of the results of `zinbwave` is the low-dimensional matrix `W`, which is stored as the `zinbwave` component of the `reducedDims` slot of the `SingleCellExperiment` object. This matrix can be retrieved using the `reducedDim()` extractor.

```{r}
W <- reducedDim(dataObj)
W[1:3,1:3]
```

In this workflow, the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`. The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance. To verify that `W` indeed captures the biological signal of interest, we display the MDS results in a scatterplot with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).

```{r mdsW, fig.cap="ZINB-WaVE: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(publishedClusters)], main = "",
     pch = 20, xlab = "Component 1", ylab = "Component 2")
legend(x = "topleft", legend = unique(names(col_clus)), cex = .5, 
       fill = unique(col_clus), title = "Sample")
```

@@

The next step of the workflow is to cluster the cells according to the low-dimensional matrix `W` computed in the previous step. We use the resampling-based sequential ensemble clustering (RSEC) framework implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`. Specifically, given a set of user-supplied base clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings, with the option of resampling cells and using a sequential tight clustering procedure as in [@Tseng2005]. A consensus clustering is obtained based on the levels of co-clustering of samples across the candidate clusterings. The consensus clustering is further condensed by merging similar clusters, which is done by creating a hierarchy of clusters, working up the tree, and testing for differential expression between sister nodes, with nodes of insufficient DE collapsed. As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters. For more details on the `RSEC` method and on the `clusterExperiment` package, see [@Risso2018].

Here, we are applying `RSEC` to the low-dimensional `W` matrix from ZINB-WaVE. To do so, we need to specify the `reducedDim` slot to use with the `reduceMethod` option. Here, we skip the merging step to collapse similar clusters based on the amount of differential gene expression between them. In larger dataset, this step may prevent results characterized by a large number of very small clusters (see [@Risso2018] for details).

@@

DR: Is there a way to check the older results? Where do I find the saved objects?

@@

@@

@@

The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figure \@ref(fig:addPublishedClusters) and \@ref(fig:addPublishedClusters2)), the main difference being that several of the published clusters were merged here into single clusters. This discrepancy is likely caused by the fact that we started with the top 1,000 genes, which might not be enough to discriminate between closely related clusters.

@@

@@

@@

```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clustering."}
plotReducedDims(dataObj, whichCluster="primary", reducedDim="zinbwave", pch=20,
	xlab = "Component1", ylab = "Component2", legendTitle="Sample", main="",
	plotUnassigned=FALSE
)
```

@@

We now demonstrate how to use the Bioconductor package `slingshot` to infer branching cell lineages and order cells by developmental progression along each lineage. The method, proposed in [@Street2017], comprises two main steps: (1) The inference of the global lineage structure (i.e., the number of lineages and where they branch) using a minimum spanning tree (MST) on the clusters identified above by `RSEC` and (2) the inference of cell pseudotime variables along each lineage using a novel method of simultaneous principal curves. The approach in (1) allows the identification of any number of novel lineages, while also accommodating the use of domain-specific knowledge to supervise parts of the tree (e.g., known terminal states); the approach in (2) yields robust pseudotimes for smooth, branching lineages.

This analysis is performed by the `slingshot` function and the results are stored in a `SlingshotDataSet` object. The minimal input to this function is a low-dimensional representation of the cells and a set of cluster labels; these can be separate objects (ie. a matrix and a vector) or, as below, components of a `SingleCellExperiment` object. When a `SingleCellExperiment` object is provided as input, the ouput will be an updated `SingleCellExperiment` object containing a `SlingshotDataSet` as an element of the `int_metadata` list, which can be accessed through the `SlingshotDataSet` function. For more low-level control of the lineage inference procedure, the two steps of fitting the MST and the simultaneous principal curves may be run separately via the functions `getLineages` and `getCurves`.

@@

@@

@@

```{r}
pseudoCe <- dataObj[,!primaryClusterNamed(dataObj) %in% c("-1")]
X <- reducedDim(pseudoCe,type="zinbwave")
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
lineages <- slingshot(mds$points, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1")
```

Before discussing the simultaneous principal curves, we examine the global structure of the lineages by plotting the MST on the clusters. This shows that our implementation has recovered the lineages found in the published work (Figure \@ref(fig:tree)). The `slingshot` package also includes functionality for 3-dimensional visualization as in Figure \@ref(fig:stemcelldiff), using the `plot3d` function from the package `rgl`.

```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with connecting lines between cluster centers representing the inferred global lineage structure."}
colorCl<-convertClusterLegend(pseudoCe,whichCluster="primary",output="matrixColors")[,1]
pairs(lineages, type="lineages", col = colorCl)
```

Having found the global lineage structure, `slingshot` then constructed a set of smooth, branching curves in order to infer the corresponding pseudotime variables. Simultaneous principal curves are constructed from the individual cells along each lineage, rather than the cell clusters. During this iterative process, a cell may even be reassigned to a different lineage if it is significantly closer to the corresopnding curve. This makes `slingshot` less reliant on the original clustering and generally more stable. The final curves are shown in Figure \@ref(fig:curves).

```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with smooth curves representing each inferred lineage."}
pairs(lineages, type="curves", col = colorCl)
```

@@

As an alternative, we could have incorporated the MDS results into the `pseudoCe` object and applied `slingshot` directly to it. Here, we need to specify that we want to use the MDS results, because `slingshot` would otherwise use the first element of the `reducedDims` list (in this case, the 10-dimensional `W` matrix from `zinbwave`).

```{r sling_sce}
reducedDim(pseudoCe, "MDS") <- mds$points
pseudoCe <- slingshot(pseudoCe, reducedDim = "MDS", start.clus = "c1")
pseudoCe
colData(pseudoCe)[,grep("^sling", colnames(colData(pseudoCe)))]
```

The result of `slingshot` applied to a `ClusterExperiment` object is still of class `ClusterExperiment`. Note that we did not specify a set of cluster labels, implying that `slingshot` should use the default `primaryClusterNamed` vector.

In the workflow, we recover the right ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. `slingshot` has the option for the user to provide known end cluster(s). Here is the code to use `slingshot` in a supervised setting, where we know that clusters `c3` and `c7` represent terminal cell fates.

```{r slingshotsupervised, eval=FALSE}
pseudoCeSup <- slingshot(pseudoCe, reducedDim = "MDS", start.clus = "c1",
                          end.clus = c("c3", "c6", "c2"))
```

@@

@@

DR: here, describe the transformData and plotHeatmap function. Above, probably right after RSEC, explain the ClusterExperiment class and its relation to SCE.

## Differential expression between clusters

DR: I think that it will be nice to showcase the getBestFeatures function here as well as our observational weights.

@@

DR: Not sure if we need this section. The previous future developments are now done (nice!). Perhaps here we should talk about scalability and HDF5? Kelly, is there anything you want to add here on DE along lineages?

@@

DR: Update grant information!
